---
title: "Quality control"
author: "Nils Eling"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_depth: 2
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

```{r}
#BiocManager::install("SingleCellExperiment")
library(SingleCellExperiment)
library(ggplot2)
library(ggridges)
```


# Read in the data

This part will load the single cell experiment object which was created under 1_read_in_data.

```{r load-data}
sce <- readRDS(file = "Z:/server_homes/daniels/Git/IMCWorkflow/data/extdata/sce.rds")
```

The next chunk will perform different transformations of the data.
We will first create a number of transformations (asinh co-factor 1; log10 with a pseudocount of 0.01;).
should we also add a quantile normalization, scaling?

```{r, transformation}
# asinh transformation
assay(sce, "counts") <- counts(sce)*2^16
assay(sce, "ashcounts") <- asinh(counts(sce)*2^16)
assay(sce, "log10counts") <- log10((counts(sce)*2^16)+0.01)

```

# Density plots of all markers for each transformation

This section aims at identifiying the best tranformation for your data.
Therefore we will look at the distributions of each marker intensity to identify potential global shifts and drop outs.

```{r QC, fig.width=15, fig.height=15}
# counts for one channel for each image separate
x <- data.frame(t(counts(sce)),Image = colData(sce)$image_number)
ggplot(x)+
  facet_wrap(~Image)+
  geom_density(aes(x=Intensity_MeanIntensity_FullStackFiltered_c15))
# Cell size per image
# Cell size is not reported

# Individual marker counts
library(reshape2)
dat.melt <- melt(counts(sce)) 
dat.melt$image <- sub("_.*$", "", dat.melt$Var2)
dat.melt$Indication <- sce_all$Indication[match(dat.melt$image, sce_all$image_number)]
dat.melt$ROI <- sce_all$ROI[match(dat.melt$image, sce_all$image_number)]

# Non-transformed
suppressMessages(print(ggplot(dat.melt) +
  geom_density_ridges(aes(value, Indication, fill = ROI), alpha = 0.5) + facet_wrap( ~ Var1) +
  ggtitle("Marker distribution, non-transformed") + theme(text = element_text(size = 20))))

# Arcsinh
suppressMessages(print(ggplot(dat.melt) +
  geom_density_ridges(aes(asinh(value), Indication, fill = ROI), alpha = 0.5) + facet_wrap( ~ Var1) +
  ggtitle("Marker distribution, asinh cofactor 1") + theme(text = element_text(size = 20))))

```

# Dimensionality reduction and Visualization of all cells 

We will compute a tsne and UMAP to visualize all cells.

```{r visualisation}
# tSNE
library(Rtsne)
library(RColorBrewer)
library(viridis)
set.seed(12345)
tsne <- Rtsne(t(asinh(counts(sce_all)[!grepl("HistoneH3|DNA", rownames(sce_all)),])), perplexity = 100)
reducedDims(sce_all)$TSNE <- tsne$Y

# Colour palette
colour_vec <- c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#FDBF6F", "#FF7F00")
names(colour_vec) <- unique(interaction(sce_all$Indication, sce_all$ROI))

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  Indication = sce_all$Indication,
                  ROI = sce_all$ROI)) +
  geom_point(aes(tsne1, tsne2, colour = interaction(Indication, ROI)), size = 0.1, alpha = 0.5) +
  ggtitle("log2-transformed + 0.1") + scale_color_manual(values = colour_vec)

```

```{r tsne-representation, fig.width=20, fig.height=15}
suppressPackageStartupMessages(library(cowplot))
p.list <- list()
for(i in 1:nrow(sce_all)){
  gene <- rownames(sce_all)[i]
  # plot "negative" cells first
  cur_data <- data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  marker = asinh(counts(sce_all)[gene,]))
  p.list[[i]] <- ggplot() +
  geom_point(data = cur_data[cur_data$marker <= 1,], 
             aes(tsne1, tsne2, colour = marker), size = 0.1) + 
      geom_point(data = cur_data[cur_data$marker > 1,], 
             aes(tsne1, tsne2, colour = marker), size = 0.1) +
  scale_color_viridis_c(name = gene) + 
    theme(axis.line=element_blank(),axis.text.x=element_blank(),
          axis.text.y=element_blank(),axis.ticks=element_blank(),
          axis.title.x=element_blank(),
          axis.title.y=element_blank(),
          panel.background=element_blank(),panel.border=element_blank(),panel.grid.major=element_blank(),
          panel.grid.minor=element_blank(),plot.background=element_blank())
}

plot_grid(plotlist = p.list, ncol = 7)
```


# Summary plots to identify potential images to exclude downstream

This section will plot the number of cells per image...

```{r, summary plots}

```


Next, I will visualize possible biases in ion counts across the cells.

```{r cell-specific-biases}
# Total ion count
ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  total_ion_count = colSums(counts(sce_all)))) +
  geom_point(aes(tsne1, tsne2, colour = total_ion_count), size = 0.1, alpha = 0.5) +
  scale_colour_gradientn(colours = magma(100))

# Shannon Entropy
# This measure indicates how the ion counts are distributed across the cells
prop <- t(t(counts(sce_all))/colSums(counts(sce_all)))
prod <- prop*log2(prop)
prod[is.na(prod)] <- 0
entropy <- -colSums(prod)

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  entropy = entropy)) +
  geom_point(aes(tsne1, tsne2, colour = entropy), size = 0.1, alpha = 0.5) +
  scale_colour_gradientn(colours = magma(100))
```

# Batch normalization

Here, I will observe how the batch correction affects the visualization of cells.
As per test experiments, I observed that a range based normalization using the arcsinh-transformed counts works best.

```{r batch-correction, fig.width=15, fig.height=15}
suppressPackageStartupMessages(library(cydar))

# Generate input format for cydar
# I assume that there are not batch effects between different ROIs
cydar.list <- list()
for(i in unique(sce_all$Indication)){
  cydar.list[[i]] <- list()
  cydar.list[[i]][["Y1"]] <- t(asinh(counts(sce_all))[,sce_all$Indication == i])
}

# Perform batch correction
suppressWarnings(corrected <- normalizeBatch(cydar.list, batch.comp = NULL, mode = "range"))
corrected_counts <- do.call("rbind", unlist(corrected, recursive = FALSE))

# Visualize batch effects
library(reshape2)
dat.melt <- melt(corrected_counts) 
dat.melt$image <- sub("_.*$", "", dat.melt$Var1)
dat.melt$Indication <- sce_all$Indication[match(dat.melt$image, sce_all$image_number)]
dat.melt$ROI <- sce_all$ROI[match(dat.melt$image, sce_all$image_number)]

suppressMessages(print(ggplot(dat.melt) +
  geom_density_ridges(aes(value, Indication, fill = ROI), alpha = 0.5) + facet_wrap( ~ Var2) +
  ggtitle("Marker distribution, batch-corrected") + theme(text = element_text(size = 20))))
```

```{r tsne-visualization}
# In form of tsne
set.seed(12345)
tsne <- Rtsne(corrected_counts[,!grepl("HistoneH3|DNA", colnames(corrected_counts))], perplexity = 100)

ggplot(data.frame(tsne1 = tsne$Y[,1],
                  tsne2 = tsne$Y[,2],
                  Indication = sce_all$Indication,
                  ROI = sce_all$ROI)) +
  geom_point(aes(tsne1, tsne2, colour = interaction(Indication, ROI)), size = 0.1, alpha = 0.5) +
  ggtitle("log2-transformed + 0.1") + scale_color_manual(values = colour_vec)
```

For visualization purposes, the batch-correction does not change anything.

# Save SCE object

Finally, for easy accessability, I will write the SCE object out.

```{r save-RDS}
saveRDS(sce_all, "Data/sce.rds")
```

